Viral Sequence Sorting Tools Evaluation

### 2022-09-13 TestingPipeline/Scripts/rule_tuning_2022.09.06.Rmd
New visualizations of testing sets.

Virsorter2
    Viral >= 50: +0.5
    Viral >= 0.95: +1
    Hallmark > 2: +1

@@

DeepVirFinder:
    Score >= 0.7: +0.5

@@

CheckV
    viral_genes == 0 and host_genes >= 1: keep_score = 0
    If 3*viral_genes <= host_genes: keep_score = 0
    If length > 50,000 and hallmark == 0: keep_score = 0
    If %unknown >= 75: +0.5

@@

```{r getting_viral_set_1}
getting_viral_set_1 <- function(input_seqs,
                                include_vibrant=FALSE, 
                                include_virsorter2=FALSE,
                                include_deepvirfinder=FALSE,
                                include_checkV=FALSE,
                                include_kaiju=FALSE,
                                include_virsorter=FALSE) {

@@

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus=="Yes"] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus=="Yes"] + 0.5
  }

  if (include_virsorter2) {
    keep_score[input_seqs$viral>=50] <- keep_score[input_seqs$viral>=50] + 0.5
    keep_score[input_seqs$hallmark>2] <- keep_score[input_seqs$hallmark>2] + 1
    keep_score[input_seqs$viral>=95] <- keep_score[input_seqs$viral>=95] + 1
  }

@@

  if (include_deepvirfinder) {
    keep_score[input_seqs$score>=0.7 & input_seqs$contig_length<10000] <- keep_score[input_seqs$score>=0.7 & input_seqs$contig_length<10000] + 0.5
   keep_score[input_seqs$score>=0.9 & input_seqs$contig_length<10000] <- keep_score[input_seqs$score>=0.9 & input_seqs$contig_length<10000] + 1
  }

@@

  if (include_checkV) {
    keep_score[input_seqs$percent_unknown>=75] <- keep_score[input_seqs$percent_unknown>=75] + 0.5
    keep_score[input_seqs$viral_genes==0 & input_seqs$host_genes>=1] <- 0
    keep_score[(input_seqs$viral_genes*3) <= input_seqs$host_genes] <- 0
    keep_score[input_seqs$contig_length>50000 & input_seqs$hallmark==0] <- 0
  }

@@

  precision <- table(confusion_matrix)[4]/(table(confusion_matrix)[4]+table(confusion_matrix)[2])

  recall <- table(confusion_matrix)[4]/(table(confusion_matrix)[4]+table(confusion_matrix)[1])

@@

  performance <- c(precision, recall, F1)
  names(performance) <- c("precision", "recall", "F1")

@@

this function builds a list of all of the combinations that the user wants to 
test. 
In this case, we're comparing the performance of all unique combinations of the 
six tools.
```{r}
build_score_list <- function(input_seqs, combos) {
  output <- data.frame(precision=rep(0, nrow(combos)),
                       recall=rep(0, nrow(combos)),
                       F1=rep(0, nrow(combos)))
  for (i in 1:nrow(combos)) {
    keep_score <- getting_viral_set_1(input_seqs, include_vibrant = combos$VIBRANT[i],
                                            include_virsorter = combos$VS[i],
                                            include_virsorter2 = combos$VS2[i],
                                            include_checkV = combos$CheckV[i],
                                            include_kaiju = combos$Kaiju[i],
                                            include_deepvirfinder = combos$DVF[i])

    output[i,1:3] <- assess_performance(input_seqs$seqtype, keep_score)

@@

accuracy_scores <- data.frame(testing_set_index=rep(0, nrow(combos_list)*10),
                      precision=rep(0, nrow(combos_list)*10),
                       recall=rep(0, nrow(combos_list)*10),
                       F1=rep(0, nrow(combos_list)*10))

@@

```{r}
accuracy_scores$numtools <- str_count(accuracy_scores$toolcombo, "1")
accuracy_scores <- accuracy_scores[order(accuracy_scores$numtools, decreasing=F),]
accuracy_scores$toolcombo <- factor(accuracy_scores$toolcombo, levels = unique(accuracy_scores$toolcombo))
accuracy_scores$numtools <- as.factor(accuracy_scores$numtools)
```

## Visualize how the precision, recall, and F1 scores change across pipelines.
```{r}
pal <- ggthemes::tableau_color_pal(palette="Tableau 10", type="regular")
p2 <- ggplot(accuracy_scores, aes(x=toolcombo, y=F1, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("F1 Score")
p2
ggplot(accuracy_scores, aes(x=toolcombo, y=precision)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Precision")
ggplot(accuracy_scores, aes(x=toolcombo, y=recall)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Recall")
```

>>>>

Virsorter2
    Viral >= 50: +0.5
    Viral >= 0.95: +1

@@

DeepVirFinder:
    Score >= 0.7: +0.5
    Score >= 0.9: +0.5

@@

CheckV
    If %unknown >= 75: +0.5
    Hallmark > 2: +1
    viral_genes == 0 and host_genes >= 1: keep_score = 0
    If 3*viral_genes <= host_genes: keep_score = 0
    If length > 50,000 and hallmark == 0: keep_score = 0

@@

```{r getting_viral_set_1}
getting_viral_set_1 <- function(input_seqs,
                                include_vibrant=FALSE, 
                                include_virsorter2=FALSE,
                                include_deepvirfinder=FALSE,
                                include_tuning=FALSE,
                                include_kaiju=FALSE,
                                include_virsorter=FALSE) {

@@

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus=="Yes"] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus=="Yes"] + 0.5
#    keep_score[input_seqs$vibrant_quality=="low quality draft"] <- keep_score[input_seqs$vibrant_quality=="low quality draft"] + 0.5
  }

  if (include_virsorter2) {
    keep_score[input_seqs$viral>=50] <- keep_score[input_seqs$viral>=50] + 0.5
    keep_score[input_seqs$viral>=95] <- keep_score[input_seqs$viral>=95] + 0.5
  }

@@

  if (include_deepvirfinder) {
    keep_score[input_seqs$score>=0.7 & input_seqs$contig_length<20000] <- keep_score[input_seqs$score>=0.7 & input_seqs$contig_length<20000] + 0.5
   keep_score[input_seqs$score>=0.9 & input_seqs$contig_length<20000] <- keep_score[input_seqs$score>=0.9 & input_seqs$contig_length<20000] + 0.5
  }

@@

  if (include_tuning) {
    keep_score[input_seqs$hallmark>2] <- keep_score[input_seqs$hallmark>2] + 1
    keep_score[input_seqs$percent_unknown>=75] <- keep_score[input_seqs$percent_unknown>=75] + 0.5
    keep_score[input_seqs$percent_viral>=50] <- keep_score[input_seqs$percent_viral>=50] + 0.5
    keep_score[input_seqs$viral_genes==0 & input_seqs$host_genes>=1] <- 0
    keep_score[(input_seqs$viral_genes*3) <= input_seqs$host_genes] <- 0
    keep_score[input_seqs$contig_length>50000 & input_seqs$hallmark==0] <- 0
  }

@@

  TP <- table(confusion_matrix)[4]
  FP <- table(confusion_matrix)[2]
  TN <- table(confusion_matrix)[3]
  FN <- table(confusion_matrix)[1]

  precision <- TP/(TP+FP)

  recall <- TP/(TP+FN)

@@

  MCC <- (TP*TN-FP*FN)/sqrt(as.numeric(TP+FP)*as.numeric(TP+FN)*as.numeric(TN+FP)*as.numeric(TN+FN))

  auc <- round(auc(truepositive, keep_score),4)

  #by type metrics
  fungal_FP <- table(confusion_matrix[seqtype=="fungi"])[2]
  bacterial_FP <- table(confusion_matrix[seqtype=="bacteria"])[2]
  viral_FN <- table(confusion_matrix[seqtype=="virus"])[1]

  performance <- c(precision, recall, F1, MCC, auc, fungal_FP, bacterial_FP, viral_FN)
  names(performance) <- c("precision", "recall", "F1", "MCC", "AUC", "fungal_FP",
                          "bacterial_FP", "viral_FN")

@@

this function builds a list of all of the combinations that the user wants to 
test. 
In this case, we're comparing the performance of all unique combinations of the 
six tools.
```{r}
build_score_list <- function(input_seqs, combos) {
  output <- data.frame(precision=rep(0, nrow(combos)),
                       recall=rep(0, nrow(combos)),
                       F1=rep(0, nrow(combos)),
                       MCC=rep(0, nrow(combos)),
                       AUC=rep(0, nrow(combos)),
                       fungal_FP=rep(0, nrow(combos)),
                       bacterial_FP=rep(0, nrow(combos)),
                       viral_FN=rep(0, nrow(combos)))
  for (i in 1:nrow(combos)) {
    keep_score <- getting_viral_set_1(input_seqs, include_vibrant = combos$VIBRANT[i],
                                            include_virsorter = combos$VS[i],
                                            include_virsorter2 = combos$VS2[i],
                                            include_tuning = combos$CheckV[i],
                                            include_kaiju = combos$Kaiju[i],
                                            include_deepvirfinder = combos$DVF[i])

    output[i,1:8] <- assess_performance(input_seqs$seqtype, keep_score)

@@

accuracy_scores <- data.frame(testing_set_index=rep(0, nrow(combos_list)*10),
                      precision=rep(0, nrow(combos_list)*10),
                       recall=rep(0, nrow(combos_list)*10),
                       F1=rep(0, nrow(combos_list)*10),
                       MCC=rep(0, nrow(combos_list)*10), 
                      AUC=rep(0, nrow(combos_list)*10),
                      fungal_FP=rep(0, nrow(combos_list)*10),
                      bacterial_FP=rep(0, nrow(combos_list)*10),
                      viral_FN=rep(0, nrow(combos_list)*10))

@@

```{r}
accuracy_scores$numtools <- str_count(accuracy_scores$toolcombo, "1")
#accuracy_scores <- accuracy_scores[order(accuracy_scores$numtools, decreasing=F),]
accuracy_scores <- accuracy_scores[order(accuracy_scores$MCC, decreasing=F),]
accuracy_scores$toolcombo <- factor(accuracy_scores$toolcombo, levels = unique(accuracy_scores$toolcombo))
accuracy_scores$numtools <- as.factor(accuracy_scores$numtools)
```

## Visualize how the precision, recall, and F1 scores change across pipelines.
```{r}
pal <- ggthemes::tableau_color_pal(palette="Tableau 10", type="regular")
p2 <- ggplot(accuracy_scores, aes(x=toolcombo, y=F1, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("F1 Score")
p2
ggplot(accuracy_scores, aes(x=toolcombo, y=precision, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Precision")
ggplot(accuracy_scores, aes(x=toolcombo, y=recall, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Recall")

ggplot(accuracy_scores, aes(x=precision, y=recall, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Precision") +
  ylab("Recall")

ggplot(accuracy_scores, aes(x=toolcombo, y=abs(precision-recall), 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Precision-Recall")

ggplot(accuracy_scores, aes(x=toolcombo, y=MCC, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("MCC")

ggplot(accuracy_scores, aes(x=toolcombo, y=AUC, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("AUC")

ggplot(accuracy_scores, aes(x=toolcombo, y=fungal_FP, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Fungal False Positives")

ggplot(accuracy_scores, aes(x=toolcombo, y=bacterial_FP, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Bacterial False Positives")

ggplot(accuracy_scores, aes(x=toolcombo, y=viral_FN, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (CV, DVF, KJ, VB, VS, VS2)") +
  ylab("Viral False Negatives")

# Experimenting

```{r}
viruses$keep_score <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_kaiju = T,
                                              include_tuning = T,
                                              include_virsorter = T)
```

```{r}
viruses$confusion_matrix <- "true negative"
viruses$confusion_matrix[viruses$seqtype=="virus" & viruses$keep_score<=1] <- "false negative"
viruses$confusion_matrix[viruses$seqtype=="virus" & viruses$keep_score>=1] <- "true positive"
viruses$confusion_matrix[viruses$seqtype!="virus" & viruses$keep_score>=1] <- "false positive"
```

visualizing confusion matrix by taxa
```{r}
confusion_by_taxa <- melt(table(viruses$confusion_matrix, viruses$seqtype, viruses$Index))
colnames(confusion_by_taxa) <- c("confusion_matrix", "seqtype","Index", "count")
```

```{r}
pal <- ggthemes::tableau_color_pal(palette="Tableau 10", type="regular")
```

```{r}
p2 <- ggplot(confusion_by_taxa, aes(x=count, y=as.factor(Index),
                   fill=confusion_matrix,
                   color=confusion_matrix)) +
  geom_bar(stat="identity") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(4)), 0.5),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(4)), 1),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  xlab("Number of Sequences") +
  ylab("") + 
  facet_wrap(~seqtype, scales = "free") +
  coord_flip()

p2
```
note: are these fungal actually false negatives? At least one was classified as viral by Kaiju.

# Visualizing confusion matrix by number of tools

```{r}
viruses$keep_score_visualize <- viruses$keep_score
viruses$keep_score_visualize[viruses$keep_score>1] <- ">1"

viruses$keep_score_visualize <- factor(viruses$keep_score_visualize, 
                                       levels=c("-0.5", "-1", "0", "0.5","1", ">1"))
viruses$keep_score_visualize <- factor(viruses$keep_score_visualize, 
                                       labels=c("≤ 0", "≤ 0", "≤ 0", "0.5","1", "> 1"))
```

```{r}
levels(factor(viruses$keep_score_visualize))
```

```{r}
pal <- ggthemes::tableau_color_pal(palette="Tableau 20", type="regular")

p1 <- ggplot(viruses, aes(x=as.factor(Index),
                   fill=keep_score_visualize, color=keep_score_visualize)) +
  geom_bar(stat="count", position="stack") +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16)
  ) +
  scale_color_manual(name = 'Viral Score',
                     values = alpha(c(pal(4)), 1)) +
  scale_fill_manual(name = 'Viral Score',
                     values = alpha(c(pal(4)), 0.5)) +
  xlab("Index") +
  ylab("Sequence Count") +
  facet_wrap(~confusion_matrix, scales = "free")

p1
```

# ROC 
```{r}
library(pROC)
```

```{r}
viruses$truepositive <- rep(0, nrow(viruses))
viruses$truepositive[viruses$seqtype=="virus"] <- 1
```

```{r}
rocobj <- roc(viruses$truepositive, viruses$keep_score)
rocobj_all <- roc(viruses$truepositive, viruses$keep_score_all)

auc <- round(auc(viruses$truepositive, viruses$keep_score),4)
auc_all <- round(auc(viruses$truepositive, viruses$keep_score_all),4)

#create ROC plot
ggroc(rocobj, colour = 'steelblue', size = 2) +
  ggtitle(paste0('ROC Curve ', '(AUC = ', auc, ')')) +
  coord_equal()

ggroc(rocobj_all, colour = 'green', size = 2) +
  ggtitle(paste0('ROC Curve ', '(AUC = ', auc_all, ')'))
```
Sensitivity: The probability that the model predicts a positive outcome for an observation when indeed the outcome is positive.
Specificity: The probability that the model predicts a negative outcome for an observation when indeed the outcome is negative.
